Sparse spectral and $p$-finite element methods for partial differential equations on disk-slices and trapeziums

### 2019-05-09 papers/halfdisk/halfdisk.tex
Removed SO notes and commented out stuff
comments from Xu

We show that partial differential equations become sparse linear systems when viewed as acting on expansions involving a family of orthogonal polynomials (OPs) that  generalise Jacobi polynomials, mirroring the ultraspherical spectral method for ordinary differential equations \cite{olver2013fast} and its analogue on the triangle \cite{olver2018recurrence,olver2019triangle}.  On the half-disk the family of weights we consider are of the form
$$
W^{(a,b)}(x,y) = x^a (1-x-y)^b, \qfor 0 \leq x \leq 1,\quad -\sqrt{1-x^2} \leq y \leq \sqrt{1-x^2},
$$
with corresponding OPs denoted $\hdopnkab(x,y)$, where $n$ denotes the polynomial degree, and $0 \le k \le n$. We define these to be orthogonalised lexicographically, that is,
$$
\hdopnkab(x,y) = C_{n,k} x^{n-k} y^k + (\hbox{lower order terms})
$$
where $C_{n,k} \neq 0$ and ``lower order terms'' can  include degree $n$ polynomials with fewer $y$ terms.

@@

@@

For disk slices and trapeziums, we can specialise the framework from Proposition \ref{prop:construction} in the following definition.
\begin{definition}\label{def:construction}
Let
\begin{align*}
	\Omega := \{(x,y) \in \R^2 \quad | \quad \alpha < x < \beta, \: \gamma \rho(x) < y < \delta \rho(x)\}
\end{align*}
with $\rho : (\alpha, \beta) \to \R$. Further, let $\genjacw^{(a,b,c)}(x)$ and $w^{(a,b)}_P(x)$ be two weight functions on the intervals $(\alpha, \beta)$ and $(\gamma, \delta)$ respectively, given by:
\begin{align*}
\begin{cases}
\genjacw^{(a,b,c)}(x) &:= (\beta - x)^a \: (x - \alpha)^{b} \: \rho(x)^{c} \\
w^{(a,b)}_P(x) &:= (\delta-x)^a \: (x - \gamma)^b
\end{cases}
\end{align*}
where the three-parameter family of orthogonal polynomials on $[\alpha,\beta]$ given by $\{\genjac_n^{(a,b,c)}\}$ are orthogonal with respect to $\genjacw^{(a,b,c)}(x)$, and the two-parameter family of orthogonal polynomials on $[\gamma,\delta]$ given by $\{P_n^{(a,b)}\}$ are orthogonal with respect to $w^{(a,b)}_P(x)$.

@@

For the half disk, the weight $W^{(a,b)}(x,y) = x^a \: (1-x^2-y^2)^b$ results from setting:
\begin{align*}
\begin{cases}
(\alpha,\beta) &:= (0.1) \\
(\gamma,\delta) &:= (-1,1) \\
\rho(x) &:= (1-x^2)^{\half} \\
w_1(x) &:= \genjacw^{(a,b)}(x) := x^a \: (1-x^2)^b \\
w_2(x) &:= w_P^{(b)}(x) := (1-x^2)^b = (1-x)^b \: (1+x)^b,
\end{cases}
\end{align*}
Note here the we have made the adjustment that $\genjacw^{(a,b,c)}(x) = (\beta - x)^a \: (x - \alpha)^{b} \: \rho(x)^{2c}$ and simply set the first parameter to zero and removed it for the specific half-disk case. This is for ease of writing and computation, by ensuring we reduce the need to incorporate unnecessary parameters and instead yield a two-parameter family of 2D orthogonal polynomials.

@@

We can express the three-term recurrences associated to $\genjac_n^{(a,b)}$ and $P_n^{(b,b)}$ as
\begin{align}
x \genjac_n^{(a,b)}(x) &= \beta_n^{(a,b)} \genjac_{n+1}^{(a,b)}(x) + \alpha_n^{(a,b)} \genjac_n^{(a,b)}(x) + \beta_{n-1}^{(a,b)} \genjac_{n-1}^{(a,b)}(x) 
\label{eqn:Hrecurrence} \\
y P_n^{(b,b)}(y) &= \delta_n^{(b)} P_{n+1}^{(b,b)}(y) + \gamma_n^{(b)} P_n^{(b,b)}(y) + \delta_{n-1}^{(b)} P_{n-1}^{(b,b)}(y),
\end{align}
where we note that \(\gamma_n^{(b)} = 0\), for \( n = 0,1,2,\dots\). We can use these to determine the 2D recurrences for $\hdopnkab(x,y)$:

@@

$J_x^{(a,b)}, J_y^{(a,b)}$ have banded-block-banded structures. A block-banded matrix is a block $n$-diagonal matrix where $n$ is reasonably small for the size of the full matrix. A banded-block-banded matrix is one where the blocks themselves are banded matrices. $J_x^{(a,b)}, J_y^{(a,b)}$ are given by:
\begin{align*}
J_{x/y}^{(a,b)} &= \begin{pmatrix}
		B^{x/y}_0 & A^{x/y}_0 & & & & \\
		C^{x/y}_1 & B^{x/y}_1 & A^{x/y}_1 & & & \\
		& C^{x/y}_2 & B^{x/y}_2 & A^{x/y}_2  & & & \\
		& & C^{x/y}_3 & \ddots & \ddots & \\
		& & & \ddots & \ddots & \ddots \\
	\end{pmatrix}
\end{align*}
where
\begin{align*}
A^x_n &:= \begin{pmatrix}
		\alphaab_{n+1,0,1} & 0 & \hdots & 0 \\
		& \ddots & & \vdots & \\
		& & \alphaab_{n+1,n,1} & 0 \\
	    \end{pmatrix} \in \R^{(n+1)\times(n+2)}, \quad n = 0,1,2,\dots \\
B^x_n &:= \begin{pmatrix}
		\alphaab_{n,0,2} & & \\
		& \ddots & \\
		& & \alphaab_{n,n,2} \\
	    \end{pmatrix} \in \R^{(n+1)\times(n+1)} \quad n = 0,1,2,\dots \\
C^x_n &:= \big( A^x_n \big)^\top \in \R^{(n+1)\times n},  \quad n = 1,2,\dots \\ 
\nonumber
\end{align*}
\begin{align*}
A^y_n &:= \begin{pmatrix}
		0 & \betaab_{n,0,6} & & & \\
		\betaab_{n,1,5} & 0 & & & \\
		& \ddots & \ddots & \ddots & \\
		& & \betaab_{n,n,5}& 0 & \betaab_{n,n,6} \\
	    \end{pmatrix} \in \R^{(n+1)\times(n+2)}, \quad n = 0,1,2,\dots \\
B^y_n &:= \begin{pmatrix}
		0 & \betaab_{n,0,4} & & \\
		\betaab_{n,1,3} & 0 & \ddots & \\
		& \ddots & \ddots & \betaab_{n,n-1,4} \\
		& & \betaab_{n,n,3} & 0
	    \end{pmatrix} \in \R^{(n+1)\times(n+1)}  \quad n = 0,1,2,\dots \\
C^y_n &:= \begin{pmatrix}
		0 & \betaab_{n,0,2} & & \\
		\betaab_{n,1,1} & 0 & \ddots & \\
		& \ddots & \ddots & \betaab_{n,n-2,2} \\
		& & \ddots & 0 \\
		& & & \betaab_{n,n,1} \\
	    \end{pmatrix} \in \R^{(n+1)\times n}, \quad n = 1,2,\dots
\end{align*}

@@

To obtain the recurrence coefficients for the $\{\genjac_n^{(a,b)}\}$ OPs in (\ref{eqn:Hrecurrence}), we use a version of the Stieltjes algorithm. Recall that
\begin{align*}
	x \genjac_n^{(a,b)}(x) &= \beta_n^{(a,b)} \genjac_{n+1}^{(a,b)}(x) + \alpha_n^{(a,b)} \genjac_n^{(a,b)}(x) + \beta_{n-1}^{(a,b)} \genjac_{n-1}^{(a,b)}(x), \quad n = 0,1,2,\dots
\end{align*}
where we define $\beta_{-1}^{(a,b)} = 0$ and $H_{-1}^{(a,b)}(x) \equiv 0$. We use a variant of the Stieltjes procedure \cite{gautschi1982generating} where the polynomials are expressed as Chebyshev polynomial expansions and the inner products are calculated via Clenshaw--Curtis quadrature. This has the benefit that it is easier to incorporate high-precision arithmetic, which we use to overcome ill-conditioning present when $b$ is large, as required for large $n$ in \eqref{eq:diskpolys}. The ApproxFun.jl \cite{ApproxFun} package gives a convenient way to manipulate Chebyshev and Jacobi expansions, which we use to calculate the inner products and norms in this algorithm.

@@

%
%\section{Hemisphere}
%
%We can use the similar idea of projecting up from the 2D orthogonal polynomials on the half disk to 3D OPs on the hemisphere. Define $\rho(x) := (1-x^2)^\half$ for $x \in [0,1]$ and let the half disk be denoted by $\omega$, that is let
%\begin{align}
%	\omega := \{(x,y) \in \R^2 \quad | \quad 1 < x < 0, \: \rho(x) < y < \rho(x)\}.
%\end{align}
%Further, define the hemisphere by
%\begin{align}
%	\Omega := \{(x,y,z) \in \R^3 \quad | \quad (x,y) \in \omega, \: x^2 + y^2 + z^2 = 1\}.
%\end{align}
%We can define the 3D OP family on the hemisphere as the set of polynomials $\bigQab(x,y,z)$ where
%\begin{align}
%	\bigQab_n(x,y,z) := \begin{pmatrix}
%		\bighdopab_n(x,y) \\
%		z \: \bighdop^{(a,b+1)}_{n-1}(x,y)
%	\end{pmatrix} \in \R^{2n+1}, 
%	\quad \quad 
%	\bigQab := \begin{pmatrix}
%		\bigQab_0 \\
%		\bigQab_1 \\
%		\vdots \\
%		\bigQab_N
%	\end{pmatrix}.
%\end{align}
%The relations for multiplication by $x, y$ are then trivial. For multiplication by $z$, note that $z^2 = W^{(0,1)}(x,y)$ for $(x,y,z) \in \Omega$ and consider that (as the 2D OPs for any parameters form a 2D basis of $\omega$):
%\begin{align}
%	z \: \hdopnkab(x,y) &= \sum_{m,j} c_{n,k,m,j}^{(a,b)\to(a,b+1)} \: z \: P^{(a,b+1)}_{m,j}(x,y), \\
%	z^2 \: \hdopnkab(x,y) &= \sum_{m,j} d_{n,k,m,j}^{(a,b)\to(a,b-1)} \: z \: P^{(a,b-1)}_{m,j}(x,y).
%\end{align}
%A simple calculation shows that the expansions are sparse, that is that most coefficients $c_{n,k,m,j}^{(a,b)\to(a,b+1)}$ and $d_{n,k,m,j}^{(a,b)\to(a,b-1)}$ are $0$, and are also given by
%\begin{align}
%	c_{n,k,m,j}^{(a,b)\to(a,b+1)} &= \frac{\ip< \hdopnkab, \: P^{(a,b+1)}_{m,j} >_{W^{(a,b+1)}}}{\norm{P^{(a,b+1)}_{m,j}}^2_{W^{(a,b+1)}}} \\
%	d_{n,k,m,j}^{(a,b)\to(a,b-1)} &= \frac{\ip< W^{(0,1)} \hdopnkab, \: P^{(a,b-1)}_{m,j} >_{W^{(a,b-1)}}}{\norm{P^{(a,b-1)}_{m,j}}^2_{W^{(a,b-1)}}} \nonumber \\
%	&= \frac{\ip< \hdopnkab, \: P^{(a,b-1)}_{m,j} >_{W^{(a,b)}}}{\norm{P^{(a,b-1)}_{m,j}}^2_{W^{(a,b-1)}}} \nonumber \\
%	&= \frac{\norm{\hdopnkab}_\Wab^2}{\norm{P^{(a,b-1)}_{m,j}}^2_{W^{(a,b-1)}}} \: c_{m,j,n,k}^{(a,b-1)\to(a,b)}.
%\end{align}
%The coefficients $c_{n,k,m,j}^{(a,b)\to(a,b+1)}$ and $d_{n,k,m,j}^{(a,b)\to(a,b-1)}$ are just the entries to the parameter tranformation operator matrices
%\begin{align}
%	\bighdop^{(a,b)}(x,y) &= \Big(T^{(a,b)\to(a,b+1)} \Big)^T \: \bighdop^{(a,b+1)}(x,y) \\
%	W^{(0,1)}(x,y) \bighdop^{(a,b)}(x,y) &= \Big(T^{(a,b)\to(a,b-1)} \Big)^T \: \bighdop^{(a,b-1)}(x,y)
%\end{align}
%respectively.

\sotodoinline{More Xu references.}
\sotodoinline{Add missing references to Bib file}
\bibliography{halfdisk-bib}

>>>>

We show that partial differential equations become sparse linear systems when viewed as acting on expansions involving a family of orthogonal polynomials (OPs) that  generalise Jacobi polynomials, mirroring the ultraspherical spectral method for ordinary differential equations \cite{olver2013fast} and its analogue on the triangle \cite{olver2018recurrence,olver2019triangle}.  On the half-disk the family of weights we consider are of the form
$$
W^{(a,b)}(x,y) = x^a (1-x^2-y^2)^b, \qfor 0 \leq x \leq 1,\quad -\sqrt{1-x^2} \leq y \leq \sqrt{1-x^2},
$$
with corresponding OPs denoted $\hdopnkab(x,y)$, where $n$ denotes the polynomial degree, and $0 \le k \le n$. We define these to be orthogonalised lexicographically, that is, \sotodo{Are these meant to be orthonormal?}
$$
\hdopnkab(x,y) = C_{n,k} x^{n-k} y^k + (\hbox{lower order terms})
$$
where $C_{n,k} \neq 0$ and ``lower order terms'' can  include degree $n$ polynomials with fewer $y$ terms.

@@

\sotodoinline{Mention that this is a computational challenge}

\sotodoinline{Make clear this is more than plugging in the basis into a Petrov--Galerkin method.}

@@

For disk slices and trapeziums, we can specialise the framework from Proposition \ref{prop:construction} in the following definition.
\begin{definition}\label{def:construction}
Let
\begin{align*}
	\Omega := \{(x,y) \in \R^2 \quad | \quad \alpha < x < \beta, \: \gamma \rho(x) < y < \delta \rho(x)\}
\end{align*}
with $\rho : (\alpha, \beta) \to \R$. Further, let $\genjacw^{(a,b,c)}(x)$ and $w^{(a,b)}_P(x)$ be two weight functions on the intervals $(\alpha, \beta)$ and $(\gamma, \delta)$ respectively, given by:
\begin{align*}
\begin{cases}
\genjacw^{(a,b,c)}(x) &:= (\beta - x)^a \: (x - \alpha)^{b} \: \rho(x)^{c} \\
w^{(a,b)}_P(x) &:= (\delta-x)^a \: (x - \gamma)^b
\end{cases}
\end{align*}
where the three-parameter family of orthogonal polynomials on $[\alpha,\beta]$ given by $\{\genjac_n^{(a,b,c)}\}$ are \sotodo{Make clear whether they are orthonormal or not} orthogonal with respect to $\genjacw^{(a,b,c)}(x)$, and the two-parameter family of orthogonal polynomials on $[\gamma,\delta]$ given by $\{P_n^{(a,b)}\}$ are orthogonal with respect to $w^{(a,b)}_P(x)$.

@@

For the half disk, the weight $W^{(a,b)}(x,y) = x^a \: (1-x^2-y^2)^b$ results from setting:
\begin{align*}
\begin{cases}
(\alpha,\beta) &:= (0,1) \\
(\gamma,\delta) &:= (-1,1) \\
\rho(x) &:= (1-x^2)^{\half} \\
w_1(x) &:= \genjacw^{(a,b)}(x) := x^a \: (1-x^2)^b \\
w_2(x) &:= w_P^{(b)}(x) := (1-x^2)^b = (1-x)^b \: (1+x)^b,
\end{cases}
\end{align*}
Note here the we have made the adjustment that $\genjacw^{(a,b,c)}(x) = (\beta - x)^a \: (x - \alpha)^{b} \: \rho(x)^{2c}$ and simply set the first parameter to zero and removed it for the specific half-disk case. This is for ease of writing and computation, by ensuring we reduce the need to incorporate unnecessary parameters and instead yield a two-parameter family of 2D orthogonal polynomials.

@@

We can express the three-term recurrences associated to $\genjac_n^{(a,b)}$ and $P_n^{(b,b)}$ as \sotodo{These hold only for orthonormal, which isn't the case for $P_n$}
\begin{align}
x \genjac_n^{(a,b)}(x) &= \beta_n^{(a,b)} \genjac_{n+1}^{(a,b)}(x) + \alpha_n^{(a,b)} \genjac_n^{(a,b)}(x) + \beta_{n-1}^{(a,b)} \genjac_{n-1}^{(a,b)}(x) 
\label{eqn:Hrecurrence} \\
y P_n^{(b,b)}(y) &= \delta_n^{(b)} P_{n+1}^{(b,b)}(y) + \gamma_n^{(b)} P_n^{(b,b)}(y) + \delta_{n-1}^{(b)} P_{n-1}^{(b,b)}(y),
\end{align}
where we note that \(\gamma_n^{(b)} = 0\), for \( n = 0,1,2,\dots\). We can use these to determine the 2D recurrences for $\hdopnkab(x,y)$:

@@

$J_x^{(a,b)}, J_y^{(a,b)}$ have banded-block-banded structures. A block-banded matrix is a block $n$-diagonal matrix where $n$ is reasonably small for the size of the full matrix. A banded-block-banded matrix is one where the blocks themselves are banded matrices. $J_x^{(a,b)}, J_y^{(a,b)}$ are given by:
\begin{align*}
J_{x/y}^{(a,b)} &= \begin{pmatrix}
		B^{x/y}_0 & A^{x/y}_0 & & & & \\
		C^{x/y}_1 & B^{x/y}_1 & A^{x/y}_1 & & & \\
		& C^{x/y}_2 & B^{x/y}_2 & A^{x/y}_2  & & & \\
		& & C^{x/y}_3 & \ddots & \ddots & \\
		& & & \ddots & \ddots & \ddots \\
	\end{pmatrix}
\end{align*}
where \sotodo{$C^x_n := \big( A^x_n \big)^\top$ is only true if orthonormal}
\begin{align*}
A^x_n &:= \begin{pmatrix}
		\alphaab_{n+1,0,1} & 0 & \hdots & 0 \\
		& \ddots & & \vdots & \\
		& & \alphaab_{n+1,n,1} & 0 \\
	    \end{pmatrix} \in \R^{(n+1)\times(n+2)}, \quad n = 0,1,2,\dots \\
B^x_n &:= \begin{pmatrix}
		\alphaab_{n,0,2} & & \\
		& \ddots & \\
		& & \alphaab_{n,n,2} \\
	    \end{pmatrix} \in \R^{(n+1)\times(n+1)} \quad n = 0,1,2,\dots \\
C^x_n &:= \big( A^x_n \big)^\top \in \R^{(n+1)\times n},  \quad n = 1,2,\dots \\ 
\nonumber
\end{align*}
\begin{align*}
A^y_n &:= \begin{pmatrix}
		0 & \betaab_{n,0,6} & & & \\
		\betaab_{n,1,5} & 0 & & & \\
		& \ddots & \ddots & \ddots & \\
		& & \betaab_{n,n,5}& 0 & \betaab_{n,n,6} \\
	    \end{pmatrix} \in \R^{(n+1)\times(n+2)}, \quad n = 0,1,2,\dots \\
B^y_n &:= \begin{pmatrix}
		0 & \betaab_{n,0,4} & & \\
		\betaab_{n,1,3} & 0 & \ddots & \\
		& \ddots & \ddots & \betaab_{n,n-1,4} \\
		& & \betaab_{n,n,3} & 0
	    \end{pmatrix} \in \R^{(n+1)\times(n+1)}  \quad n = 0,1,2,\dots \\
C^y_n &:= \begin{pmatrix}
		0 & \betaab_{n,0,2} & & \\
		\betaab_{n,1,1} & 0 & \ddots & \\
		& \ddots & \ddots & \betaab_{n,n-2,2} \\
		& & \ddots & 0 \\
		& & & \betaab_{n,n,1} \\
	    \end{pmatrix} \in \R^{(n+1)\times n}, \quad n = 1,2,\dots
\end{align*}

@@

To obtain the recurrence coefficients for the $\{\genjac_n^{(a,b)}\}$ OPs in (\ref{eqn:Hrecurrence}), we use a version of the Stieltjes algorithm. Recall that\sotodo{check this}
\begin{align*}
	x \genjac_n^{(a,b)}(x) &= \beta_n^{(a,b)} \genjac_{n+1}^{(a,b)}(x) + \alpha_n^{(a,b)} \genjac_n^{(a,b)}(x) + \beta_{n-1}^{(a,b)} \genjac_{n-1}^{(a,b)}(x), \quad n = 0,1,2,\dots
\end{align*}
where we define $\beta_{-1}^{(a,b)} = 0$ and $H_{-1}^{(a,b)}(x) \equiv 0$. We use a variant of the Stieltjes procedure \cite{gautschi1982generating} where the polynomials are expressed as Chebyshev polynomial expansions and the inner products are calculated via Clenshaw--Curtis quadrature. This has the benefit that it is easier to incorporate high-precision arithmetic, which we use to overcome ill-conditioning present when $b$ is large, as required for large $n$ in \eqref{eq:diskpolys}. The ApproxFun.jl \cite{ApproxFun} package gives a convenient way to manipulate Chebyshev and Jacobi expansions, which we use to calculate the inner products and norms in this algorithm.

@@

\bibliography{halfdisk-bib}
